HTW EXAM Fits and Transfer Predictions

### 2023-11-13 Simulations/htw_exam.qmd
model tables

```{r}
#| label: fig-alm-diagram
#| fig.cap: The basic structure of the ALM model.

alm_plot()

@@

@@

::: {.column-page-inset-right}
## Model Table
### ALM Activation & Response
| Step | Equation | Description |
|------|----------|-------------|
| **ALM Activation & Response**  |           |
| Input Activation | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$ | Activation of each input node $X_i$, is a function of the Gaussian similarity between the node value and stimulus X.  |
| Output Activation                 | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$ | Activation of each Output unit $O_j$ is the weighted sum of the input activations and association weights.   |
| Output Probability                | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$ | Each output node has associated response, $Y_j$. The probability of response $Y_j$ is determined by the ratio of output activations.  |
| Mean Output                       | $m(x) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$   | The response to stimulus x is the weighted average of the response probabilities. |
| **ALM Learning**                  |                                                                                                                                                          |
| Feedback Activation               | $f_j(Z) = e^{-c(Z-Y_j)^2}$                  | After responding, feedback signal Z is presented, activating each output node via the Gaussian similarity to the ideal response. |
| Update Weights                    | $w_{ji}(t + 1) = w_{ji}(t) + \alpha \cdot (f_j(Z(t)) - O_j(X(t)) \cdot a_i(X(t))$    | Delta rule to update weights. Magnitude of weight changes controlled by learning rate parameter alpha.  |  
| **EXAM**                          |                                                                                                                                                          |
| Extrapolation                     | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$ | Novel test stimulus X activates input nodes associated with trained stimuli.     |
|                                   | $E[Y|X_i] = m(X_i) + \frac{m(X_{i+1})-m(X_{i-1})}{X_{i+1}-X_{i-1}} \cdot [X - X_i]$ | Slope value computed from nearest training instances and then added to the response associated with the nearest training instance,m(x)    |               
: ALM & EXAM Equations {#tbl-alm-exam}
:::

Following the procedure used by @mcdanielPredictingTransferPerformance2009, we will assess the ability of both ALM and EXAM to account for the empirical data when fitting the models to 1) only the training data, and 2) both training and testing data. Models will be fit directly to the trial by trial data of each individual participants, both by minimizing the root-mean squared deviation (RMSE), and by maximizing log likelihood. Because ALM has been shown to do poorly at accounting for human patterns extrapolation [@deloshExtrapolationSineQua1997], we will also fit the extended EXAM version of the model, which operates identically to ALM during training, but includes a linear extrapolation mechanism for generating novel responses during testing.

For the hybrid model, predictions are computed by first generating separate predictions from ALM and EXAM, and then combining them using the following equation: $\hat{y} = (1 - w) \cdot alm_pred + w \cdot exam_pred$. For the grid search, the weight parameter is varied from 0 to 1, and the resulting RMSE is recorded.

```{r}

almParamsV <- cbind(Model="ALM Test Only",pluck(a_te_v, "Fit"), pluck(a_te_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)) ) |>
  rbind(cbind(Model="ALM Test & Train", pluck(a_tetr_v,"Fit"), pluck(a_tetr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="ALM Train Only", pluck(a_tr_v, "Fit"), pluck(a_tr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

examParamsV <- cbind(Model="EXAM Test Only",pluck(ex_te_v, "Fit"), pluck(ex_te_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="EXAM Test & Train", pluck(ex_tetr_v,"Fit"), pluck(ex_tetr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="EXAM Train Only", pluck(ex_tr_v, "Fit"), pluck(ex_tr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

hybridParamsV <-cbind(Model="Hybrid Test Only",pluck(hybrid_te_v, "Fit"), pluck(hybrid_te_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="Hybrid Test & Train", pluck(hybrid_tetr_v,"Fit"), pluck(hybrid_tetr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="Hybrid Train Only", pluck(hybrid_tr_v, "Fit"), pluck(hybrid_tr_v, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

almParamsC <- cbind(Model="ALM Test Only",pluck(a_te_c, "Fit"), pluck(a_te_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)) ) |>
  rbind(cbind(Model="ALM Test & Train", pluck(a_tetr_c,"Fit"), pluck(a_tetr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="ALM Train Only", pluck(a_tr_c, "Fit"), pluck(a_tr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

examParamsC <- cbind(Model="EXAM Test Only",pluck(ex0_te_c, "Fit"), pluck(ex0_te_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="EXAM Test & Train", pluck(ex0_tetr_c,"Fit"), pluck(ex0_tetr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="EXAM Train Only", pluck(ex0_tr_c, "Fit"), pluck(ex0_tr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

hybridParamsC <-cbind(Model="Hybrid Test Only",pluck(hybrid_te_c, "Fit"), pluck(hybrid_te_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred))) |>
  rbind(cbind(Model="Hybrid Test & Train", pluck(hybrid_tetr_c,"Fit"), pluck(hybrid_tetr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  rbind(cbind(Model="Hybrid Train Only", pluck(hybrid_tr_c, "Fit"), pluck(hybrid_tr_c, "test") %>% summarise(Test_RMSE=RMSE(y,pred)))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

@@

```{r}
#| label: tbl-e1-model-fits2V
#| tbl-cap: Varied Group - Fit Parameters and Model RMSE

pander(almParamsV, caption="ALM"); pander(examParamsV, caption="EXAM"); pander(hybridParamsV,caption="Hybrid") 
```

```{r}
#| label: tbl-e1-model-fitsC
#| tbl-cap: Constant Group - Fit Parameters and Model RMSE

pander(almParamsC, caption="ALM"); pander(examParamsC, caption="EXAM"); pander(hybridParamsC,caption="Hybrid")
```

@@

>>>>

```{r}
#| label: fig-alm-diagram
#| fig.cap: The basic structure of the ALM model. 
 alm_plot()
```

@@

Following the procedure used by @mcdanielPredictingTransferPerformance2009, we will assess the ability of both ALM and EXAM to account for the empirical data when fitting the models to 1) only the training data, and 2) both training and testing data. Models were fit to the aggregated participant data by minimizing the root-mean squared deviation (RMSE). Because ALM has been shown to do poorly at accounting for human patterns extrapolation [@deloshExtrapolationSineQua1997], we will also generate predictions from the EXAM model for the testing stage. EXAM which operates identically to ALM during training, but includes a linear extrapolation mechanism for generating novel responses during testing.

For the hybrid model, predictions are computed by first generating separate predictions from ALM and EXAM, and then combining them using the following equation: $\hat{y} = (1 - w) \cdot alm.pred + w \cdot exam.pred$. For the grid search, the weight parameter is varied from 0 to 1, and the resulting RMSE is recorded.

@@

```{r}
#| label: tbl-e1-cogmodel
#| tbl-cap: Fit Parameters and Model RMSE. The Test_RMSE column is the main performance indicator of interest, and represents the RMSE for just the testing data. The Fit_Method column indicates the data used to fit the model.
#| column: screen-inset-right
#|

create_combined_df <- function(model_names, model_data, group) {
  do.call(rbind, Map(function(name, data) {
    model<- ifelse(grepl("Hybrid", name), "Hybrid", ifelse(grepl("EXAM", name), "EXAM", "ALM"))
    fit_method <- gsub(".*(Test Only|Test & Train|Train Only).*", "\\1", name)
    extract_params(model, data, group, fit_method)
  }, model_names, model_data))
}

# Adjust the extract_params function to accept and include the fit_method parameter
extract_params <- function(model_name, model_data, group, fit_method) {
  params_df <- cbind(Model = model_name, Group = group, Fit_Method = fit_method,
                     pluck(model_data, "Fit"), pluck(model_data, "test") %>% summarise(Test_RMSE = round(RMSE(y, pred)),1)) %>% 
                     mutate(across(where(is.numeric), round, 4))
  if (!"w" %in% names(params_df)) {
    params_df$w <- NA
  }
  return(params_df)
}

model_classes <- c("ALM", "EXAM", "Hybrid")
groups <- c("V", "C")

model_names_v <- c("ALM Test Only", "ALM Test & Train", "ALM Train Only", "EXAM Test Only", "EXAM Test & Train", "EXAM Train Only", "Hybrid Test Only", "Hybrid Test & Train", "Hybrid Train Only")

#model_names_v <- c("Test Only", "Test & Train", "Train Only")
model_names_c <- model_names_v

model_data_v <- list(a_te_v, a_tetr_v, a_tr_v, ex_te_v, ex_tetr_v, ex_tr_v, hybrid_te_v, hybrid_tetr_v, hybrid_tr_v)
model_data_c <- list(a_te_c, a_tetr_c, a_tr_c, ex0_te_c, ex0_tetr_c, ex0_tr_c, hybrid_te_c, hybrid_tetr_c, hybrid_tr_c)

combined_params_v <- create_combined_df(model_names_v, model_data_v, "Varied")
combined_params_c <- create_combined_df(model_names_c, model_data_c, "Constant")

all_combined_params <- rbind(combined_params_v, combined_params_c)

library(flextable)

identity <- function (x){x}
fmt_iden <- function(x, digits = 2) {
  paste0(x)
}

# all_combined_params %>% 
#   group_by(Group, Model) |> summarise(across(c(c,lr,Test_RMSE), ~max(.x,na.rm=TRUE)))

grouped_params <- all_combined_params %>% 
  group_by(Group, Model) %>%
  summarise(across(c(c, lr, w, Value, Test_RMSE), ~identity(.x))) %>%
  ungroup()

ft <- flextable(all_combined_params) %>% 
  theme_vanilla() %>% 
  align(align = "left", part = "all") %>% 
  colformat_double(digits = 2) %>% 
  autofit()

# Pre-calculate the averages and standard deviations
all_combined_params2 <- all_combined_params %>%
  group_by(Fit_Method, Group, Model) %>%
  summarise(
    c_stats = fmt_iden(c),
    lr_stats = fmt_iden(lr),
     rmse_stats = fmt_iden(Test_RMSE),
    .groups = "drop"
  )

# Use the pre-calculated stats in tabulator
tab <- tabulator(
  x = all_combined_params2, rows = c("Fit_Method", "Model"),
  columns = "Group",
  `c` = as_paragraph(c_stats),
  `lr` = as_paragraph(lr_stats),
  `Test_RMSE` = as_paragraph(rmse_stats)
) %>%
  as_flextable() |> align(align="left",part="all")

tab

# a=aggregate(c ~ Group+Model+Fit_Method, data=all_combined_params,identity)
# 
# all_combined_params |> group_by(Model, Group,Fit_Method) %>% 
#   summarise(
#     across(
#       all_of(c("c", "lr")),
#       list(
#         c = ~ identity(.x),
#         lr = ~ identity(.x)
#       )
#     )
#   ) %>% flextable() %>% separate_header()
# 
# 
# ft_1 <-  tabulator(
#   x = all_combined_params,
#   rows = "Group",
#   columns = c("Fit_Method","Model"),
#   `c1` = c,
#   `$lr$` = as_paragraph(lr) ) %>% as_flextable()
# 
# ft_1
```

```{r}
#| label: tbl-e1-hybridParam
#| tbl-cap: Hybrid Model - w parameter. The $w$ parameter determines the balance between the ALM and EXAM response generation processes, and is only included for the hybrid model. A weight of .5 would indicate equal contribution from both models. $w$ values approaching 1 indicate stronger weight for EXAM.

extract_params <- function(model_name, model_data) {
  cbind(Model = model_name,
        pluck(model_data, "Fit"),
        pluck(model_data, "test") %>% summarise(Test_RMSE = RMSE(y, pred))
       ) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)))
}

# Define model names and data
# model_names <- c("ALM Test Only", "ALM Test & Train", "ALM Train Only", "EXAM Test Only", "EXAM Test & Train", "EXAM Train Only", "Hybrid Test Only", "Hybrid Test & Train", "Hybrid Train Only")
# model_data_v <- list(a_te_v, a_tetr_v, a_tr_v, ex_te_v, ex_tetr_v, ex_tr_v, hybrid_te_v, hybrid_tetr_v, hybrid_tr_v)
# model_data_c <- list(a_te_c, a_tetr_c, a_tr_c, ex0_te_c, ex0_tetr_c, ex0_tr_c, hybrid_te_c, hybrid_tetr_c, hybrid_tr_c)

# params_list_v <- Map(extract_params, model_names, model_data_v)
# params_list_c <- Map(extract_params, model_names, model_data_c)

# almParamsV <- do.call(rbind, params_list_v[1:3])
# examParamsV <- do.call(rbind, params_list_v[4:6])
# hybridParamsV <- do.call(rbind, params_list_v[7:9])

# almParamsC <- do.call(rbind, params_list_c[1:3])
# examParamsC <- do.call(rbind, params_list_c[4:6])
# hybridParamsC <- do.call(rbind, params_list_c[7:9])

all_combined_params |> filter(Model =="Hybrid") |> 
  group_by(Group, Fit_Method) |>
  summarise(w=fmt_iden(first(w)), Test_RMSE=(fmt_iden(Test_RMSE))) |>
  group_by(Group,Fit_Method) |> flextable()

@@

@@

::: {.column-page-inset-right}
## Model Table
### ALM Activation & Response
| Step | Equation | Description |
|------|----------|-------------|
| **ALM Activation & Response**  |           |
| Input Activation | $a_i(X) = \frac{e^{-c(X-X_i)^2}}{\sum_{k=1}^M e^{-c(X-X_k)^2}}$ | Activation of each input node $X_i$, is a function of the Gaussian similarity between the node value and stimulus X.  |
| Output Activation                 | $O_j(X) = \sum_{k=1}^M w_{ji} \cdot a_i(X)$ | Activation of each Output unit $O_j$ is the weighted sum of the input activations and association weights.   |
| Output Probability                | $P[Y_j|X] = \frac{O_j(X)}{\sum_{k=1}^M O_k(X)}$ | Each output node has associated response, $Y_j$. The probability of response $Y_j$ is determined by the ratio of output activations.  |
| Mean Output                       | $m(x) = \sum_{j=1}^L Y_j \cdot \frac{O_j(x)}{\sum_{k=1}^M O_k(X)}$   | The response to stimulus x is the weighted average of the response probabilities. |
| **ALM Learning**                  |                                                                                                                                                          |
| Feedback Activation               | $f_j(Z) = e^{-c(Z-Y_j)^2}$                  | After responding, feedback signal Z is presented, activating each output node via the Gaussian similarity to the ideal response. |
| Update Weights                    | $w_{ji}^{new}=w_{ji}+\eta\alpha_{ji}$   | Delta rule to update weights. Magnitude of weight changes controlled by learning rate parameter $\alpha$.  |  
| **EXAM**                          |                                                                                                                                                          |
| Extrapolation                     | $P[X_i|X] = \frac{a_i(X)}{\sum_{k=1}^M a_k(X)}$ | Novel test stimulus X activates input nodes associated with trained stimuli.     |
|                                   | $E[Y|X_i] = m(X_i) + \frac{m(X_{i+1})-m(X_{i-1})}{X_{i+1}-X_{i-1}} \cdot [X - X_i]$ | Slope value computed from nearest training instances and then added to the response associated with the nearest training instance,m(x)    |               
: ALM & EXAM Equations {#tbl-alm-exam}
:::

### 2023-11-13 _working/_htw_exam.qmd
model tables

Each model was fit to the data in 3 different ways. 1) To just the testing data, 2) Both the training and testing data, 3) Only the training data. In all cases, the model only updates its weights during the training phase, and the weights are frozen during the testing phase. In all cases, only the ALM model generates predictions during the training phase. For the testing phase, all 3 models are used to generate predictions.

The Value column of the table reflects the minimum RMSE achieved when fitting the model. The Test_RMSE column is the main performance indicator of interest, and represents the RMSE for just the testing data - thus for the Test Only fits, the Value and Test_RMSE columns are identical - given that the models were trained and tested on the exact some data sets.  For the Test & Train, and Train Only fits - the Test_RMSE can diverge from the fitting values.

>>>>

Each model was fit to the data in 3 different ways. 1) To just the testing data, 2) Both the training and testing data, 3) Only the training data. In all cases, the model only updates its weights during the training phase, and the weights are frozen during the testing phase. In all cases, only the ALM model generates predictions during the training phase. For the testing phase, all 3 models are used to generate predictions. The models' performance under each fitting scenario is detailed in Table 1, showcasing parameters such as the generalization parameter (c), learning rate (lr), and the RMSE for both constant and varied training conditions.
